Properties of Winning Iterated Prisoner's Dilemma Strategies.

### 2023-12-30 paper/main.tex
move last sections around
addressing MH last comment

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades, from the celebrated performance of Tit for Tat to the
introduction of the zero-determinant strategies and the use of sophisticated learning
structures such as neural networks. Many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a small number of tournaments, casting doubt on the generalizability
of conclusions. In this work, we analyze a large collection of \numberofstrategies
strategies in thousands of computer tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features.
The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios,
nevertheless there are several properties that heavily influence the best performing
strategies. This refines the properties described by Axelrod in light of
recent and more diverse opponent populations to: be nice, be provocable and generous,
be a little envious, be clever, and adapt to the environment. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities.
The features of high performing strategies help cast some light on why
strategies such as Tit For Tat performed historically well in tournaments and
why zero-determinant strategies typically do not fare well in tournament
settings. Furthermore, our findings have implications for the future training of
autonomous agents, as understanding the crucial features for incorporation into
these agents becomes essential.
\end{abstract}

@@

It is important to note that we do not assert the existence of a single
best-performing strategy across all tournaments or tournament types. On the
contrary, our work demonstrates that such a strategy does not exist
(notwithstanding a few strategies with broadly high performance). The primary
objective of this paper, presented in the latter parts of the paper, is to
continue the discussion on the properties of successful strategies, a
conversation started by Axelrod. 
The results of our analysis conclude that the properties of a successful
strategy in the Iterated Prisoner's Dilemma (\IPD{}) are:

@@

This manuscript explores the performance of \numberofstrategies strategies in
the \IPD{} in thousands of computer tournaments. The collection of computer
tournaments presented here is the largest and most diverse in the literature.
The \numberofstrategies strategies are drawn from \axelrod{} library and include
strategies from the \IPD{} literature. The computer tournaments encompass four
different types. So, what is the best way to play the \IPD{}? And is there a
single dominant strategy for the \IPD{}? There was not a single strategy within
the collection of \numberofstrategies strategies that managed to perform well in
all the tournament variations it competed in. A strategy ranking highly in a
specific environment did not guarantee its success over different tournament
types, with a few exceptions - strategies that generalize better. Already
well-known in the AI/ML literature, adding noise to training data leads to more
robust models. We see that clearly here, where the strategies trained for noise
(or designed for noise) tend to be better generalists. There were instances
where a few strategies trained in narrow conditions outperformed more generalist
strategies, as they tend to overfit. However, the strategies trained with noise
perform well in general, whilst the strategies trained specifically on no noise
or small subpopulations do not.

@@

A variety of open-source software have been used in this work. The authors would
like to express their gratitude to the open-source software community, whose
invaluable contributions significantly enhanced the development and execution of
this research. Namely, the authors would like to thank the developers of the
following software packages: Axelrod-Python library for IPD
simulations~\cite{axelrodproject}, the Matplotlib library for
visualisation~\cite{hunter2007matplotlib}, The Numpy library for data
manipulation~\cite{walt2011numpy}, and finally the scikit-learn library for data
analysis~\cite{scikit-learn}.

@@

>>>>

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades, from the celebrated performance of Tit for Tat to the
introduction of the zero-determinant strategies and the use of sophisticated learning
structures such as neural networks. Many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a small number of tournaments, casting doubt on the generalizability
of conclusions. In this work, we analyze a large collection of \numberofstrategies
strategies in thousands of computer tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. 
We find that although a strategy's high ranking in a specific environment
does not guarantee success across different tournament types, certain strategies
consistently perform well across various settings. Notably,
strategies designed or trained for noise are the strategies that generalise
better—an observation well established in AI/ML literature where adding noise to
training data enhances model robustness.
We identify several crucial properties
heavily influencing the success of strategies, refining Axelrod's described
properties in light of encounters with more diverse opponents to: being
cooperative, responsive to provocation, generous yet slightly envious, clever,
and adaptable to the prevailing environment. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities.
The features of high performing strategies help cast some light on why
strategies such as Tit For Tat performed historically well in tournaments and
why zero-determinant strategies typically do not fare well in tournament
settings. Furthermore, our findings have implications for the future training of
autonomous agents, as understanding the crucial features for incorporation into
these agents becomes essential.
\end{abstract}

@@

It is important to note that we do not assert the existence of a single
best-performing strategy across all tournaments or tournament types. On the
contrary, our work demonstrates that such a strategy does not always exist
(notwithstanding a few strategies with broadly high performance). The primary
objective of this paper, presented in the latter parts of the paper, is to
continue the discussion on the properties of successful strategies, a
conversation started by Axelrod. 
The results of our analysis conclude that the properties of a successful
strategy in the Iterated Prisoner's Dilemma (\IPD{}) are:

@@

This manuscript explores the performance of \numberofstrategies strategies in
the \IPD{} in thousands of computer tournaments. The collection of computer
tournaments presented here is the largest and most diverse in the literature.
The \numberofstrategies strategies are drawn from \axelrod{} library and include
strategies from the \IPD{} literature. The computer tournaments encompass four
different types. So, what is the best way to play the \IPD{}? And is there a
single dominant strategy for the \IPD{}?
A strategy ranking highly in a
specific environment did not guarantee its success over different tournament
types, with a few exceptions - strategies that generalize better. Already
well-known in the AI/ML literature, adding noise to training data leads to more
robust models. We see that clearly here, where the strategies trained for noise
(or designed for noise) tend to be better generalists. There were instances
where a few strategies trained in narrow conditions outperformed more generalist
strategies, as they tend to overfit. However, the strategies trained with noise
perform well in general, whilst the strategies trained specifically on no noise
or small subpopulations do not.

@@

@@

N.G. acknowledges the generous support of the European Research Council Starting
Grant 850529: E-DIRECT and the Max Planck Society.
A variety of open-source software have been used in this work. The authors would
like to express their gratitude to the open-source software community, whose
invaluable contributions significantly enhanced the development and execution of
this research. Namely, the authors would like to thank the developers of the
following software packages: Axelrod-Python library for IPD
simulations~\cite{axelrodproject}, the Matplotlib library for
visualisation~\cite{hunter2007matplotlib}, The Numpy library for data
manipulation~\cite{walt2011numpy}, and finally the scikit-learn library for data
analysis~\cite{scikit-learn}.
